### 2017-02-08 manuscript.rmd
pdf formatting, small changes to text throughout

# Multiple pairwise comparisons between species can strongly mislead functional genomics studies

@@

There is now considerable interest in comparing functional genomics data across species. One goal of this work is to provide an integrated understanding of genome and phenotype evolution. Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species. These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=5, fig.cap="Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events. As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}

	# Plot the contrasts for duplication events and speciation events
	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent contrasts" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
		theme_classic()

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

	g_phylo_null = 
		nodes_sim_null_contrast %>% 
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
			geom_freqpoly( binwidth=.0025, position="identity" ) +
			xlab( "Phylogenetic Independent Contrast" ) +
			ylab( "Density" ) +
			xlim( 0, 0.05 ) + 
			ylim( 0, 75 ) +
			theme_classic() + 
			ggtitle( "(b) Null simulation, phylogenetic" ) +
			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

	g_pair_null = 
		pairwise_sim_null_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(a) Null simulation, pairwise" ) +
			theme(legend.position="none")

	g_phylo_oc = 
	nodes_sim_oc_contrast %>% 
	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.0025, position="identity" ) +
		xlab( "Phylogenetic Independent Contrast" ) +
		ylab( "Density" ) +
		xlim( 0, 0.05 ) + 
		ylim( 0, 75 ) +
		theme_classic() + 
		ggtitle( "(d) OC simulation, phylogenetic" ) +
		theme(legend.position="none")

	g_pair_oc = 
		pairwise_sim_oc_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(c) OC simulation, pairwise" ) +
			theme(legend.position="none")

@@

To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures. First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events. Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a). As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age. Phylogenetic analysis of the data simulated under the null hypothesis (Figure 5b) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.

@@

![Distributions of pairwise similarity scores for each phase of development. Pairwise scores for the ctenophore are red. Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right. Model of variance, which is inversely related to similarity, is on the left. (a) The distributions as published. Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression. The five least-similar mid phase scores were all from the ctenophore. Published KS p-values, based on duplicated data, are in parentheses. The inset ctenophore image is by S. Haddock from phylopic.org. (b) The distributions after the exclusion of the ctenophore. The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.png?raw=true)

@@

Thanks to XX. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).

@@

```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="There is a negative relationship between Tau and the maximum observed expression for each gene."}

@@

	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
		geom_point( size=0.5, alpha = 0.05 ) + 
		geom_smooth( method="lm" ) +
		xlab( "Maximum Expression" ) +
		theme_classic()

>>>>

# Pairwise comparisons are problematic when analyzing functional genomic data across species

@@

There is now considerable interest in comparing functional genomics data across species. One goal of this work is to provide an integrated understanding of genome and phenotype evolution. Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species. The statistical problems that arise from not considering these relationships can lead pairwise approaches to the wrong conclusions, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare gene expression across species with pairwise methods. We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique features of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.height=2, fig.width=4, fig.cap="Phylogneetic reanalysis of the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw]. Density plot of the magnitude of phylogenetic independent contrasts for Tau following duplication and speciation events. As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}

	# Plot the contrasts for duplication events and speciation events
	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent contrasts" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		theme_classic() + 
		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture (OC), while the phylogenetic method provides different results under each model. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

	g_phylo_null = 
		nodes_sim_null_contrast %>% 
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
			geom_freqpoly( binwidth=.0025, position="identity" ) +
			xlab( "Phylogenetic Independent Contrast" ) +
			ylab( "Density" ) +
			xlim( 0, 0.05 ) + 
			ylim( 0, 75 ) +
			theme_classic() + 
			ggtitle( "(b) Null, phylogenetic" ) +
			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

	g_pair_null = 
		pairwise_sim_null_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(a) Null, pairwise" ) +
			theme(legend.position="none")

	g_phylo_oc = 
	nodes_sim_oc_contrast %>% 
	ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.0025, position="identity" ) +
		xlab( "Phylogenetic Independent Contrast" ) +
		ylab( "Density" ) +
		xlim( 0, 0.05 ) + 
		ylim( 0, 75 ) +
		theme_classic() + 
		ggtitle( "(d) OC, phylogenetic" ) +
		theme(legend.position="none")

	g_pair_oc = 
		pairwise_sim_oc_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(c) OC, pairwise" ) +
			theme(legend.position="none")

@@

To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures. First, we simulated the evolution of Tau under the null model that it evolves at the same rate following duplication and speciation events. Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a). As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age. Phylogenetic analysis of the data simulated under the null hypothesis (Figure 5b) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.

@@

![Distributions of pairwise similarity scores for each phase of development. Pairwise scores for the ctenophore are red. Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right. Model of variance, which is inversely related to similarity, is on the left. (a) The distributions as published. Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression. The five least-similar mid phase scores were all from the ctenophore. Published KS p-values, based on duplicated data, are in parentheses. The inset ctenophore image is by S. Haddock from phylopic.org. (b) The distributions after the exclusion of the ctenophore. The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.pdf)

@@

Thanks to Steve Haddock, Alex Damian Serrano, and August Guang for feedback on the manuscript. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).

@@

There is a negative relationship between miximum expression and Tau for each gene (Supplementary Figure 9).

```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="Genes with higher maximum observed expression tend to have lower Tau. The blue line is the linear model for the relationship between the two."}

@@

	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
		geom_point( size=0.5, alpha = 0.05 ) + 
		geom_smooth( method="lm", se=FALSE ) +
		xlab( "Maximum Expression" ) +
		ylim( 0, 1 ) +
		theme_classic()
